---
title: "Exploracion_inicial"
author: "Gabriela Centeno y Sofía Zorrilla"
format: html
knitr:
  opts_knit:
    root.dir: "../../"
---

Load libraries

```{r}
#| eval: false
library(dplyr)
library(readr)
library(data.table)
library(ggplot2)
library(here)
library(sf)
library(here)
```

-   Load raw data from GBIF and Red de Herbarios Mexicanos records of oaks in Mexico. Also load the species name matching data frame that was generated with the backbone of gbif.

```{r}
#| eval: false

#--- Load the GBIF data frame that previously was separated in three ---
files_list <- list.files(here::here("data/in/gbif_dwc/division_occurrence"), pattern=".csv", full.names=TRUE)


files <- lapply(files_list,fread)

recordsGbif_raw <- do.call(rbind,files)

# --- Load the Red de Herbarios Mexicanos data frame ---
recordsHerbario_raw <- fread(here::here("data/in/herbariomex_dw/occurrences.csv"))


```

-Compare the names from gbif and Red de Herbarios Mexicanos with the backbone of gbif, so we can know which are synonyms and which are not.

```{r}
library(rgbif)
library(readr)
library(dplyr)


# Función para consultar cada nombre en el backbone de GBIF
match_species <- function(name) {
  result <- name_backbone(name = name)
  
  if (is.null(result)) {
    # Si no encuentra resultado, devuelve NA
    return(data.frame(verbatim_name = name, usageKey = NA, canonicalName = NA, status = NA, confidence = NA))
  } else {
    # Si encuentra, devuelve los datos como data.frame
    result_df <- as.data.frame(result)
    result_df$verbatim_name <- name  # Guarda el nombre original
    return(result_df)
  }
}

# Aplicar la función a las dos bases de datos
backbone_gbif <- recordsGbif_raw %>%
  pull(scientificName) %>%                      # Extrae solo la columna species
  lapply(match_species) %>%              # Aplica la función a cada nombre
  bind_rows()                            # Combina los resultados

backbone_herb <- recordsHerbario_raw %>%
  pull(scientificName) %>%                      
  lapply(match_species) %>%              
  bind_rows()  


# Save it
write_csv(backbone_gbif, 
          here::here("data/in/backbone/gbif_nombres_backbone.csv"))

write_csv(backbone_herb, 
          here::here("data/in/backbone/herb_final_backbone.csv"))

```



-Join the GBIF and Red de Herbarios Mexicanos dataframes with their species matching data frame.

```{r}

# --- Load the species name matching data frame ---
gbif_nombres_backbone <- fread(here::here("data/in/backbone/gbif_nombres_backbone.csv"))

herb_nombres_backbone <- fread(here::here("data/in/backbone/herb_final_backbone.csv"))


# --- Add a unique id to each value in the gbif and herbario data frames--- 
recordsGbif_raw <- recordsGbif_raw %>%
  mutate(id_interno = paste0("GBIF_", row_number()))

recordsHerbario_raw <- recordsHerbario_raw %>%
  mutate(id_interno = paste0("Herb_", row_number()))


# --- Join them with their species matching data frame--- 

herb_corr_raw <- cbind(recordsHerbario_raw, herb_nombres_backbone)
gbif_corr_raw <- cbind(recordsGbif_raw, gbif_nombres_backbone)


```

-Join the GBIF and Red de Herbarios Mexicanos dataframes together. For this, it is necessary to add a suffix to each column. Also, it is necessary to make all the columns characters and select the columns of interest.

```{r}
#---Add a suffix to avoid duplicate columns--- 
names(herb_corr_raw) <- make.names(names(herb_corr_raw), unique = TRUE)
names(gbif_corr_raw) <- make.names(names(gbif_corr_raw), unique = TRUE)

#---Make all columns character and select columns of interest for both data frames--- 
herb_corr_raw <- herb_corr_raw %>%
  mutate(across(everything(), as.character)) %>% 
  select(scientificName, canonicalName, correctname, rank, decimalLatitude, decimalLongitude, stateProvince, county, municipality,locality, eventDate, year, month, recordedBy, id_interno,basisOfRecord)

gbif_corr_raw <- gbif_corr_raw %>%
  mutate(across(everything(), as.character)) %>% 
  select(scientificName, canonicalName, correctname, rank, decimalLatitude, decimalLongitude, stateProvince, county, municipality,locality, eventDate, year, month, recordedBy, id_interno,basisOfRecord)

#--- Join rows from both data frames ---
gbif_herb_raw <- bind_rows(herb_corr_raw, gbif_corr_raw)

```

-Then filter the records that have only genera and the records that are reliable
```{r}
gbif_herb_raw <- gbif_herb_raw %>% filter(!scientificName %in% c("Quercus","Quercus L.")) %>%
  filter(basisOfRecord %in% c("HUMAN_OBSERVATION", "PRESERVED_SPECIMEN", "OBSERVATION","PreservedSpecimen","Observation","HumanObservation","Preserved Specimen","LivingSpecimen","Foto","Ejemplar herborizado"))
```

Write csv

```{r}
write.csv(gbif_herb_raw, here::here("data/temp/gbif_herb_raw.csv"))
```


-Proceed to coordinate_cleaning.qmd