---
title: "coordinate_cleaning"
author: "Gabriela Centeno y Sofía Zorrilla"
format: html
knitr:
  opts_knit:
    root.dir: "../../"
---

Before this doc, you have to go to joint_tables.qmd.

Upload the necessary libraries.

```{r}
library(data.table)
library(CoordinateCleaner)
library(terra)
library(here)
library(tidyverse)
library(data.table)
library(ggplot2)
```


Upload the data frame that has the gbif and the herbario information.

```{r}
gbif_herb_raw <- fread(here::here("data/temp/gbif_herb_raw.csv"))
```


-Delete all the registers that do not have coordinates and with valid coordinates
```{r}
hgbif_herb_coor <- gbif_herb_raw %>%
  filter(!is.na(decimalLatitude) & !is.na(decimalLongitude)) %>%
    filter(decimalLongitude >= -118 & decimalLongitude <= -87,  
           decimalLatitude >= 14 & decimalLatitude <= 33)      
```

-Erase the duplicates of the registers that have the same coordinates, name of specie and location.

```{r}
hgbif_no_duplicates <- hgbif_herb_coor %>%
  distinct(decimalLatitude, decimalLongitude, correctname, .keep_all = TRUE)

```

Eliminar datos de capitales, centroides, cordenadas en el mar o instituciones

```{r}
hgbif_no_duplicates_clean <- clean_coordinates(x = hgbif_no_duplicates, 
                               lon = "decimalLongitude",
                               lat = "decimalLatitude",
                               species = "correctname", 
                               value = "clean")

#Default tests Default = c("capitals", "centroids", "equal", "gbif", "institutions", "outliers", "seas", "zeros")
```


Delete the records of the species that doesn't fullfill the requirements 
```{r}
# Vector of species to remove
species_to_remove <- c( "Lithocarpus corneus", "Lithocarpus elegans", "Lithocarpus korthalsii", "Quercus acutangula", "Quercus alpescens", "Quercus alvordiana", "Quercus augustini",  "Quercus bourgaei", "Quercus brenesii", "Quercus buckleyi", "Quercus caudicans",
  "Quercus cerris", "Quercus charcasana", "Quercus chartacea", "Quercus chiapensis",
  "Quercus copeyensis", "Quercus douglasii", "Quercus dysophylla", "Quercus edwardsiae",
  "Quercus ellipsoidalis", "Quercus floccosa", "Quercus ganderi", "Quercus geminata",
  "Quercus ghiesbreghtii", "Quercus gilva", "Quercus ilex", "Quercus juergensenii",
  "Quercus laevis", "Quercus laurifolia", "Quercus magnifolia", "Quercus manzanillana",
  "Quercus melissae", "Quercus montana", "Quercus morehus", "Quercus nigra",
  "Quercus obtusanthera", "Quercus perseaefolia", "Quercus petraea", "Quercus robur",
  "Quercus robusta", "Quercus schultzei", "Quercus serrata", "Quercus suber",
  "Quercus supranitida", "Quercus tardifolia", "Quercus undata", "Quercus undulata",
  "Quercus vallicola", "Quercus verde", "Quercus virginiana", "Quercus aculcingensis", "Roldana Candicans"
)

# Removing rows where species column matches any in species_to_remove
hgbif_clean <- hgbif_no_duplicates_clean %>%
  filter(!(correctname %in% species_to_remove))

write.csv(hgbif_clean, file = here::here("data/in/hgbif_clean.csv"))

data_summary <- hgbif_clean %>%
  group_by(correctname) %>%
  summarise(
    n_registros = n(),
    # IUCN = unique(IUCN),
    # elev_min = min(elevacion_inegi, na.rm = TRUE),
    # elev_max = max(elevacion_inegi, na.rm = TRUE),
    #seccion = first(section)
  ) %>%
  ungroup() %>%
  arrange(by = n_registros)

```


-Proceed to extract_elevation.qmd


