---
title: "SES_calculation"
author: "Gabriela Centeno y Sofía Zorrilla"
format: html
eval: false
knitr:
  opts_knit:
    root.dir: "../../"
---

*This .qmd document calculates the SES for PD and WE.*


##Libraries

-   Load the necessary libraries

```{r}
library(phyloraster)
library(terra)
library(ape)
library(phylobase)
library(here)
library(data.table)
library(here)
library(stringr)
library(ggplot2)
library(tidyverse)
library(SESraster)
```

##Stack of Rasters

-   Enter the stack of rasters of the absence/presence of the 177 species we made previously. 

```{r}
file_list <- list.files(here::here("data/out/raster_species/"), full.names = TRUE, pattern = ".tif$") # raster stacks de las diferentes escalas.

files <- lapply(file_list, terra::rast)

```

*Here we upload the df of the tree and the records data frame*

```{r}
data_morton <- fread(here::here("data/in/filogenia/Hipp2019_sample_metadata.csv"))
geo_data <- fread(here::here("data/in/hgbif_completo_iucn.csv"))
```

Now, we view the similarities and the differencies between the two.

```{r}
#TODO: Revisar si los nombres en el correctname que no coinciden con los del morton realmnete no están en la filogenia 
#TODO: Decidir si esta es la filogenia que vamos a utilizar

common_morton <- data_morton[`Cleaned_NAMES-USE-THIS` %in% geo_data$correctname,]
common_geo <- geo_data[correctname %in% common_morton$`Cleaned_NAMES-USE-THIS`]

diff<- setdiff(unique(data_morton$`Cleaned_NAMES-USE-THIS`),unique(geo_data$correctname))


unique(common_geo$correctname)
```

Existen 87 especies en común entre la filogenia y la base de datos de gbif/herbario.

Now we save all the records in the gbif/herbario data frame that are present in the tree.
```{r}
common_geo <- merge(common_geo,unique(common_morton[,.(`Cleaned_NAMES-USE-THIS`,section,clade,subgenus)]), by.x = "correctname", by.y="Cleaned_NAMES-USE-THIS",all.x = T)

fwrite(common_geo, here::here('data/out/common_geo.csv'), row.names = F)
```


## Phylogenetic tree

### Clean names of the tree


The next function allows you to replace characters at the tips of the tree.

```{r}
library(ape)
library(treeio)
library(rotl)

## --- Functions --- ##

#' Modify Tip Labels of a Phylogenetic Tree
#'
#' This function modifies the tip labels of a phylogenetic tree based on a user-defined string transformation function.
#' By default, it removes substrings matching '_ott.*' and replaces underscores with spaces.
#'
#' @param tree A phylogenetic tree (likely of class `phylo`) containing tip labels.
#' @param modify_label_fn A function that takes a character vector of labels and returns a modified character vector.
#' The default function removes '_ott.*' and replaces underscores with spaces.
#'
#' @return A modified phylogenetic tree with updated tip labels.
#' @importFrom tibble as_tibble
#' @importFrom stringr str_replace str_remove str_detect
modify_tips <- function(tree, modify_label_fn = function(x) str_remove(x, '_ott.*')) {
  
  metadata <- as_tibble(tree) %>% mutate(label = modify_label_fn(label))
  tree$tip.label <- metadata[which(str_detect(metadata$label, 'Quercus')),]$label
  
  return(tree)
}

```


El arbol en crudo muestra etiquetas para cada especie. Como se muestra en el siguiente ejemplo: Quercus chrysolepis\|USA\|CA\|CA-DAV-MH20\|QUE000293'

Para este caso, no se utilizaran las etiquetas. Por lo que se utiliza la función modify_tips para eliminar las etiquetas y eliminar los espacios, por lo que el ejemplo quedaría de la siguiente manera: Quercus_chrysolepis

```{r}
complete_tree <- read.tree(here::here('data/in/filogenia/OpenTreeOfLife_hipp2019.tre'))


complete_tree <- modify_tips(complete_tree, modify_label_fn = function(x) str_replace_all(str_remove_all(str_remove(x, '\\|.*'),"\\'"),' |-','_'))

```

Cargar las distribuciones de las especies.

Hacer un subset de los rasters que si se encuentran en el arbol (87).

```{r}
#files = rasters de distribucion por cada escala

sub_dist <- list()

for(i in seq_along(files)) {

names(files[[i]]) <- str_replace_all(names(files[[i]]),' |-','_')
rasters_to_use <- which(names(files[[i]]) %in% str_replace_all(unique(common_geo$correctname),' |-','_'))

sub_dist[[i]] <- subset(files[[i]], rasters_to_use)

}
```

```{r}
# Revisar que los nombres de las especies estén en el orden que están en el árbol

#Toma las puntas del arbol y busca las que no se encuentran en la base de datos, es decir, y con el sub_tree se eliminan
tips_to_drop <-  which(!complete_tree$tip.label %in% str_replace_all(unique(common_geo$correctname),' |-','_'))
sub_tree <- drop.tip(complete_tree,tips_to_drop)


# Con el paquete phyloraster, se reacomodan los datos de los rasters para que esten en el orden de la filogenia. 

#Esto arroja un Warning, el cual indica que las especies Quercus_depressipes, Quercus_gambelii no se encuentran en el arbol. Por lo que solo se estaria calculando la PD de 85 especies.

dataprep <- list()
pdr <- list()

for(i in seq_along(sub_dist)) {
  
  #tabla con especies ordenadas
dataprep[[i]] <- phylo.pres(x = sub_dist[[i]], tree = sub_tree)

  #el raster del calculo
pdr[[i]] <- rast.pd(x = dataprep[[i]]$x, dataprep[[i]]$tree)

}
```

## Calculate standard effective size (SES)
```{r}
x <- sub_dist[[3]]
tree <- sub_tree

# Genera la tabla ordenada
data <- phylo.pres(x, tree)


# Calcula SES
tiempo <- system.time({
t <- rast.pd.ses(data$x, 
                 edge.path = data$edge.path,
                 branch.length = data$branch.length, 
                 aleats = 999,
                 filename = here::here("data/out/SES/PD_SES_999aleats.tif"), 
                 overwrite = TRUE)
})

print(tiempo)
#     user   system  elapsed 
# 1076.109   11.529 1086.029 

# Guarda también un respaldo seguro
saveRDS(t, here::here("data/out/SES/PD_SES_999aleats.rds"))

pd_ses <- rast(here::here("data/out/SES/PD_SES_999aleats.tif"))

png(here::here("data/out/SES/ses_pd.png"), width = 1200, height = 800)
plot(pd_ses)
dev.off()

### WE

# Calcula SES
tiempo <- system.time({
t <- rast.we.ses(data$x, 
                 aleats = 999,
                 filename = here::here("data/out/SES/WE_SES_999aleats.tif"), 
                 overwrite = TRUE)
})

print(tiempo)
#     user   system  elapsed 
# 1076.109   11.529 1086.029 

# Guarda también un respaldo seguro
saveRDS(t, here::here("data/out/SES/WE_SES_999aleats.rds"))

we_ses <- rast(here::here("data/out/SES/WE_SES_999aleats.tif"))


png(here::here("data/out/SES/ses_we.png"), width = 1200, height = 800)
plot(we_ses)
dev.off()

```

```{r}
library(rnaturalearth)
library(rnaturalearthdata)
library(viridis)
library(sf)

pd_ses$SES.PD

# Get Mexico boundaries
mexico <- st_read(here::here("data/in/biogeo_provinces/all_bio_regions/pbiogmx17gw.shp"))

# Get Mexico's bounding box for raster creation
bbox <- st_bbox(mexico)

# Create a sample raster using terra (temperature gradient as example)
# You can replace this with your own raster data
r <- rast(nrows = 100, ncols = 100,
          xmin = bbox["xmin"], xmax = bbox["xmax"],
          ymin = bbox["ymin"], ymax = bbox["ymax"])



# Convert the entire stack to a dataframe at once
raster_df <- as.data.frame(pd_ses, xy = TRUE)
raster_df <- raster_df[!is.na(raster_df$SES.PD),]




# Create the plot
plot_ses <- raster_df %>% 
  filter(SES.PD != 0) %>% 
  mutate(pval = ifelse(SES.PD > 0, p_lower.PD,ifelse(SES.PD < 0, p_upper.PD,NA))) %>% 
ggplot() +
  # Plot raster layer
  
  # Add Mexico state boundaries
  geom_sf(data = mexico, fill = NA, color = "#8B8989", linewidth = 0.2) +
  geom_raster(aes(x = x, y = y, fill = SES.PD,alpha = pval < 0.05 )) +
  # Color scale
  scale_fill_viridis(option = "viridis", name = "SES") +
  scale_alpha_manual(values = c(0.2,1), guide = FALSE)+
  # Labels and theme
  labs(title = "SES Diversidad filogenética",
       caption = "Tamaños de efecto de PD, colores fríos(cálidos) representan\nsitios con menor(mayor) PD que lo esperado por la hipotesis nula.\nLas celdas con transparencia representan sitios no significativos.",
       x = "Longitud",
       y = "Latitud", alpha = "") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    legend.position = "right"
  ) +
  coord_sf()

ggsave(here::here("data/out/SES/PD_biogeographicRegions_SES_signif.png"), plot_ses, width = 20, height = 20, units = "cm", bg = "white")
```

```{r}
library(ape)

plot_highlighted_tree <- function(tree, species_to_keep, 
                                   highlight_color = "red",
                                   default_color = "black",
                                   edge_width = 2,
                                   plot_subset = FALSE,
                                   main = NULL,
                                   verbose = TRUE,
                                  cex = NULL) {
  
  # Find species that are actually in the tree
  species_in_tree <- species_to_keep[species_to_keep %in% tree$tip.label]
  species_missing <- species_to_keep[!species_to_keep %in% tree$tip.label]
  
  # Print warning if species are missing
  if (length(species_missing) > 0 && verbose) {
    warning(paste("The following species are not in the tree and will be skipped:\n",
                  paste(species_missing, collapse = ", ")))
  }
  
  # Check if we have any valid species
  if (length(species_in_tree) == 0) {
    stop("None of the specified species are in the tree!")
  }
  
  # Subset the tree with only valid species
  tree_subset <- keep.tip(tree, species_in_tree)
  
  # Plot subset tree if requested
  if (plot_subset) {
    plot(tree_subset, main = "Subset Tree")
  }
  
  # Initialize all edges with default color
  edge_colors <- rep(default_color, nrow(tree$edge))
  
  # Get node numbers for tips in the subset
  tips_to_highlight <- which(tree$tip.label %in% species_in_tree)
  
  # Color edges that lead directly to highlighted tips
  for (i in 1:nrow(tree$edge)) {
    # Check if this edge leads to one of our tips
    if (tree$edge[i, 2] %in% tips_to_highlight) {
      edge_colors[i] <- highlight_color
    }
  }
  
  # Plot the full tree with colored branches
  plot(tree, edge.color = edge_colors, edge.width = edge_width,
       main = main, cex = cex)
  
  # Return the subset tree and info (invisible)
  invisible(list(
    tree_subset = tree_subset,
    species_found = species_in_tree,
    species_missing = species_missing
  ))
}

plot_ses_map <- function(mexico_sf, 
                         pd_data,
                         x_label = "Longitud",
                         y_label = "Latitud",
                         border_color = "#8B8989",
                         border_width = 0.2,
                         highlight_color = "black",
                         viridis_option = "viridis",
                         legend_name = "SES",
                         alpha_values = c(0.2, 1)) {
  
  ggplot() +
    # Add Mexico state boundaries
    geom_sf(data = mexico_sf, fill = NA, color = border_color, 
            linewidth = border_width) +
    # Add high PD cells
    geom_tile(data = pd_data, aes(x = x, y = y), fill = highlight_color) +
    # Color scale
    scale_fill_viridis(option = viridis_option, name = legend_name) +
    scale_alpha_manual(values = alpha_values, guide = "none") +
    # Labels and theme
    labs(x = x_label,
         y = y_label,
         alpha = "") +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 16, face = "bold"),
      plot.subtitle = element_text(size = 12),
      legend.position = "right"
    ) +
    coord_sf()
}

```


```{r}

high_PD <- raster_df[raster_df$SES.PD > 4,]

highSESmap <- plot_ses_map(pd_data = high_PD, mexico_sf = mexico)

raster_species <- as.data.frame(x, xy = TRUE)
species_high_PD <- raster_species[rownames(high_PD),]
species_high_PD <- colSums(species_high_PD[,-c(1,2)]) > 0

# Extract species names from your dataframe
species_to_keep <- names(species_high_PD[species_high_PD])

# Dave pdf
pdf(here::here("data/out/SES/highSES_values.pdf"), width = 10, height = 8)
print(highSESmap)
plot_highlighted_tree(tree, species_to_keep, cex = 0.5)
title(main = "Ejemplo de celdas con SES altos (>4), significativos", line = 0.5)
dev.off()
```

```{r}

low_PD <- raster_df[raster_df$SES.PD < -2,]

lowSESmap <- plot_ses_map(pd_data = low_PD, mexico_sf = mexico)

raster_species <- as.data.frame(x, xy = TRUE)
species_low_PD <- raster_species[rownames(low_PD),]
species_low_PD <- colSums(species_low_PD[,-c(1,2)]) > 0

# Extract species names from your dataframe
species_to_keep <- names(species_low_PD[species_low_PD])

# Dave pdf
pdf(here::here("data/out/SES/lowSES_values.pdf"), width = 10, height = 8)
print(lowSESmap)
plot_highlighted_tree(tree, species_to_keep, cex = 0.5)
title(main = "Ejemplo de celdas con SES bajos (<-2), significativos", line = 0.5)
dev.off()

plot_highlighted_tree(tree, species_to_keep, main = "Ejemplo de celdas con SES bajos, significativos")

```



########## ENDEMISMO ######
```{r}

we_ses$SES.WE

# Get Mexico boundaries
mexico <- st_read(here::here("data/in/biogeo_provinces/all_bio_regions/pbiogmx17gw.shp"))

# Get Mexico's bounding box for raster creation
bbox <- st_bbox(mexico)

# Create a sample raster using terra (temperature gradient as example)
# You can replace this with your own raster data
r <- rast(nrows = 100, ncols = 100,
          xmin = bbox["xmin"], xmax = bbox["xmax"],
          ymin = bbox["ymin"], ymax = bbox["ymax"])



# Convert the entire stack to a dataframe at once
raster_df <- as.data.frame(we_ses, xy = TRUE)
raster_df <- raster_df[!is.na(raster_df$SES.WE),]




# Create the plot
plot_ses <- raster_df %>% 
  filter(SES.WE != 0) %>% 
  mutate(pval = ifelse(SES.WE > 0, p_lower.WE,ifelse(SES.WE < 0, p_upper.WE,NA))) %>% 
ggplot() +
  # Plot raster layer
  
  # Add Mexico state boundaries
  geom_sf(data = mexico, fill = NA, color = "#8B8989", linewidth = 0.2) +
  geom_raster(aes(x = x, y = y, fill = SES.WE,alpha = pval < 0.05 )) +
  # Color scale
  scale_fill_viridis(option = "viridis", name = "SES") +
  scale_alpha_manual(values = c(0.2,1), guide = FALSE)+
  # Labels and theme
  labs(title = "SES Endemismo Ponderado",
       caption = "Tamaños de efecto de WE, colores fríos(cálidos) representan\nsitios con menor(mayor) PD que lo esperado por la hipotesis nula.\nLas celdas con transparencia representan sitios no significativos.",
       x = "Longitud",
       y = "Latitud", alpha = "") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    legend.position = "right"
  ) +
  coord_sf()

ggsave(here::here("data/out/SES/WE_biogeographicRegions_SES_signif.png"), plot_ses, width = 20, height = 20, units = "cm", bg = "white")
```

```{r}
library(ape)

plot_highlighted_tree <- function(tree, species_to_keep, 
                                   highlight_color = "red",
                                   default_color = "black",
                                   edge_width = 2,
                                   plot_subset = FALSE,
                                   main = NULL,
                                   verbose = TRUE,
                                  cex = NULL) {
  
  # Find species that are actually in the tree
  species_in_tree <- species_to_keep[species_to_keep %in% tree$tip.label]
  species_missing <- species_to_keep[!species_to_keep %in% tree$tip.label]
  
  # Print warning if species are missing
  if (length(species_missing) > 0 && verbose) {
    warning(paste("The following species are not in the tree and will be skipped:\n",
                  paste(species_missing, collapse = ", ")))
  }
  
  # Check if we have any valid species
  if (length(species_in_tree) == 0) {
    stop("None of the specified species are in the tree!")
  }
  
  # Subset the tree with only valid species
  tree_subset <- keep.tip(tree, species_in_tree)
  
  # Plot subset tree if requested
  if (plot_subset) {
    plot(tree_subset, main = "Subset Tree")
  }
  
  # Initialize all edges with default color
  edge_colors <- rep(default_color, nrow(tree$edge))
  
  # Get node numbers for tips in the subset
  tips_to_highlight <- which(tree$tip.label %in% species_in_tree)
  
  # Color edges that lead directly to highlighted tips
  for (i in 1:nrow(tree$edge)) {
    # Check if this edge leads to one of our tips
    if (tree$edge[i, 2] %in% tips_to_highlight) {
      edge_colors[i] <- highlight_color
    }
  }
  
  # Plot the full tree with colored branches
  plot(tree, edge.color = edge_colors, edge.width = edge_width,
       main = main, cex = cex)
  
  # Return the subset tree and info (invisible)
  invisible(list(
    tree_subset = tree_subset,
    species_found = species_in_tree,
    species_missing = species_missing
  ))
}

plot_ses_map <- function(mexico_sf, 
                         we_data,
                         x_label = "Longitud",
                         y_label = "Latitud",
                         border_color = "#8B8989",
                         border_width = 0.2,
                         highlight_color = "black",
                         viridis_option = "viridis",
                         legend_name = "SES",
                         alpha_values = c(0.2, 1)) {
  
  ggplot() +
    # Add Mexico state boundaries
    geom_sf(data = mexico_sf, fill = NA, color = border_color, 
            linewidth = border_width) +
    # Add high PD cells
    geom_tile(data = we_data, aes(x = x, y = y), fill = highlight_color) +
    # Color scale
    scale_fill_viridis(option = viridis_option, name = legend_name) +
    scale_alpha_manual(values = alpha_values, guide = "none") +
    # Labels and theme
    labs(x = x_label,
         y = y_label,
         alpha = "") +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 16, face = "bold"),
      plot.subtitle = element_text(size = 12),
      legend.position = "right"
    ) +
    coord_sf()
}

```


```{r}

high_WE <- raster_df[raster_df$SES.WE > 12,]

highSESmap <- plot_ses_map(we_data = high_WE, mexico_sf = mexico)

raster_species <- as.data.frame(x, xy = TRUE)
species_high_WE <- raster_species[rownames(high_WE),]
species_high_WE <- colSums(species_high_WE[,-c(1,2)]) > 0

# Extract species names from your dataframe
species_to_keep <- names(species_high_WE[species_high_WE])

# Dave pdf
pdf(here::here("data/out/SES/highWE_values.pdf"), width = 10, height = 8)
print(highSESmap)
plot_highlighted_tree(tree, species_to_keep, cex = 0.5)
title(main = "Ejemplo de celdas con SES altos (>12), significativos", line = 0.5)
dev.off()
```

```{r}

low_WE <- raster_df[raster_df$SES.WE < 0,]

lowSESmap <- plot_ses_map(we_data = low_WE, mexico_sf = mexico)

raster_species <- as.data.frame(x, xy = TRUE)
species_low_WE <- raster_species[rownames(low_WE),]
species_low_WE <- colSums(species_low_WE[,-c(1,2)]) > 0

# Extract species names from your dataframe
species_to_keep <- names(species_low_WE[species_low_WE])

# Dave pdf
pdf(here::here("data/out/SES/lowWE_0_values.pdf"), width = 10, height = 8)
print(lowSESmap)
plot_highlighted_tree(tree, species_to_keep, cex = 0.5)
title(main = "Ejemplo de celdas con SES bajos (< 0), significativos", line = 0.5)
dev.off()

plot_highlighted_tree(tree, species_to_keep, main = "Ejemplo de celdas con SES bajos, significativos")

```
