---
title: "coordinate_cleaning"
author: "Gabriela Centeno y Sofía Zorrilla"
format: html
knitr:
  opts_knit:
    root.dir: "../../"
---

Before this doc, you have to go to joint_tables.qmd.

Upload the necessary libraries.

```{r}
library(data.table)
library(dplyr)
library(CoordinateCleaner)
library(terra)
library(here)
library(dplyr)
library(data.table)
library(ggplot2)
```


Upload the data frame that has the gbif and the herbario information.

```{r}
gbif_herb_raw <- fread(here::here("data/temp/gbif_herb_raw.csv"))
```


-Delete all the registers that do not have coordinates and with valid coordinates
```{r}
hgbif_herb_coor <- gbif_herb_raw %>%
  filter(!is.na(decimalLatitude) & !is.na(decimalLongitude)) %>%
    filter(decimalLongitude >= -118 & decimalLongitude <= -87,  
           decimalLatitude >= 14 & decimalLatitude <= 33)      
```

-Erase the duplicates of the registers that have the same coordinates, name of specie and location.

```{r}
hgbif_no_duplicates <- hgbif_herb_coor %>%
  distinct(decimalLatitude, decimalLongitude, correctname, .keep_all = TRUE)

```

Eliminar datos de capitales, centroides, cordenadas en el mar o instituciones

```{r}
hgbif_no_duplicates_clean <- clean_coordinates(x = hgbif_no_duplicates, 
                               lon = "decimalLongitude",
                               lat = "decimalLatitude",
                               species = "correctname", 
                               value = "clean")
```


Delete the records of the species that doesn't fullfill the requirements 
```{r}
# Vector of species to remove
species_to_remove <- c(
  "Quercus chiapensis",
  "Quercus acutangula",
  "Quercus manzanillana",
  "Quercus schultzei",
  "Quercus obtusanthera",
  "Quercus edwardsiae",
  "Quercus juergensenii",
  "Quercus vallicola",
  "Quercus perseaefolia",
  "Quercus magnifolia",
  "Quercus augustini",
  "Quercus supranitida",
  "Quercus undulata",
  "Quercus verde",
  "Quercus ignaciensis",
  "Quercus floccosa",
  "Quercus bourgaei",
  "Quercus alpescens",
  "Quercus brenesii",
  "Quercus chartacea",
  "Quercus ghiesbreghtii",
  "Quercus undata",
  "Quercus rekonis",
  "Quercus dysophylla",
  "Quercus caudicans"
)

# Removing rows where species column matches any in species_to_remove
hgbif_clean <- hgbif_no_duplicates_clean %>%
  filter(!(correctname %in% species_to_remove))

write.csv(hgbif_clean, file = here::here("data/in/hgbif_clean.csv"))

```


-Proceed to extract_elevation.qmd


