---
title: "coordinate_cleaning"
author: "Gabriela Centeno y Sofía Zorrilla"
format: html
knitr:
  opts_knit:
    root.dir: "../../"
---

Before this doc, you have to go to joint_tables.qmd.

Upload the necessary libraries.

```{r}
library(data.table)
library(dplyr)
library(CoordinateCleaner)
library(terra)
library(here)
library(dplyr)
library(data.table)
library(ggplot2)
```


Upload the data frame that has the gbif and the herbario information.

```{r}
gbif_herb_raw <- fread(here::here("data/temp/gbif_herb_raw.csv"))
```


-Delete all the registers that do not have coordinates and with valid coordinates
```{r}
hgbif_herb_coor <- gbif_herb_raw %>%
  filter(!is.na(decimalLatitude) & !is.na(decimalLongitude)) %>%
    filter(decimalLongitude >= -118 & decimalLongitude <= -87,  
           decimalLatitude >= 14 & decimalLatitude <= 33)      
```

-Erase the duplicates of the registers that have the same coordinates, name of specie and location.

```{r}
hgbif_no_duplicates <- hgbif_herb_coor %>%
  distinct(decimalLatitude, decimalLongitude, correctname, .keep_all = TRUE)

```

Eliminar datos de capitales, centroides, cordenadas en el mar o insituciones

```{r}
hgbif_no_duplicates_clean <- clean_coordinates(x = hgbif_no_duplicates, 
                               lon = "decimalLongitude",
                               lat = "decimalLatitude",
                               species = "correctname", 
                               value = "clean")
```


Delete the records of the species that doesn't fullfill the requirements 
```{r}
# Vector of species to remove
species_to_remove <- c("Quercus ilex", "Quercus tardifolia", "Quercus buckleyi", 
                       "Quercus geminata", "Quercus laurifolia", "Quercus augustini", 
                       "Quercus petraea", "Quercus gilva", "Quercus breedloveana", 
                       "Quercus robur", "Lithocarpus elegans", "Quercus alpescens", 
                       "Quercus robusta", "Quercus suber", "Quercus douglasii", 
                       "Lithocarpus korthalsii", "Quercus charcasana", "Lithocarpus corneus", 
                       "Quercus cerris", "Quercus ganderi", "Quercus ellipsoidalis", 
                       "Quercus alvordiana", "Quercus montana", "Quercus morehus", 
                       "Quercus virginiana", "Quercus laevis", "Quercus copeyensis", 
                       "Quercus nigra", "Quercus serrata")

# Removing rows where species column matches any in species_to_remove
hgbif_clean <- hgbif_clean %>%
  filter(!(correctname %in% species_to_remove))

write.csv(hgbif_clean, file = here::here("data/in/hgbif_clean.csv"))

```


-Proceed to extract_elevation.qmd


