---
title: "richness_wendemism_calculation"
author: "Gabriela Centeno y Sofía Zorrilla"
format: html
eval: false
knitr:
  opts_knit:
    root.dir: "../../"
---

*This .qmd document calculates the richness and weighted endemism for all the species using the phyloraster package.*

*Then, it calculates the phylogenetic diversity, richness and weighted endemism for the 87 species present in the phylogenetic tree.*

##Libraries

-   Load the necessary libraries

```{r}
library(phyloraster)
library(terra)
library(ape)
library(phylobase)
library(here)
library(data.table)
library(here)
library(stringr)
library(ggplot2)
library(tidyverse)
```

##Stack of Rasters

-   Enter the stack of rasters of the absence/presence of the 177 species we made previously. Then calculate the richness with function rast.sr, finally plot it.

```{r}
file_list <- list.files(here::here("data/out/raster_species/"), full.names = TRUE, pattern = ".tif")
```

##Richness Calculation

*With the rast.sr funcion of the phyloraster, we are going to calculate the richness*

```{r}

files <- lapply(file_list, terra::rast)
richness <- lapply(files, rast.sr)


plot(richness[[3]], main = "Species richness")
```

###Save the richness rasters and the images of it.

*With the next functions we are going to save all the rasters and images generated*

```{r}
#| eval: false 

#' Save a raster object to file
#'
#' This function saves a raster object to the specified file path using the `writeRaster` function.
#'
#' @param output_path Character. The file path (relative to project root) where the raster will be saved.
#' @param raster A Raster* object to be saved.
#'
#' @return No return value. Called for its side effect of saving the raster file.
#' @export
save_raster <- function(output_path, raster) {
  writeRaster(raster, here::here(output_path), overwrite = TRUE)
}

#' Save a raster plot as PNG
#'
#' This function saves a visual plot of a raster object as a PNG image with a specified title.
#'
#' @param output_path Character. The file path (relative to project root) where the PNG will be saved.
#' @param raster A Raster* object to be plotted and saved.
#' @param title Character. The title to be displayed on the raster plot.
#'
#' @return No return value. Called for its side effect of saving the PNG image.
#' @export
save_png <- function(output_path, raster, title) {
  png(here::here(output_path), width = 3000, height = 2000, res = 300)
  plot(raster, main = title)
  dev.off()
}

```

```{r}
outpath_tif <- file.path('data/out/richness_tif',c(paste0('richness_tif_',c(0.135, 0.225, 0.315, 0.405, 0.450),'.tif')))

outpath_png <- file.path('data/out/richness_png',c(paste0('richness_png_',c(0.135, 0.225, 0.315, 0.405, 0.450),'.png')))

titles <- c(paste0('Riqueza de especies (escala ',c(0.135, 0.225, 0.315, 0.405, 0.450), '°)'))



lapply(seq_along(richness), function(x){
  save_raster(raster = richness[[x]], output_path =  outpath_tif[[x]])
  })

lapply(seq_along(richness), function(x){
  save_png(raster = richness[[x]], output_path =  outpath_png[[x]], title = titles[[x]])
  })

```

##Weighted endemism calculation

*We calculated the weighted endemism with the function rast.we, then plot it.*

```{r}

wendemism <- list()


for(i in seq_along(files)) {

  
wendemism[[i]] <- rast.we(x = files[[i]])

}

plot(wendemism[[5]]$WE, main = "Weighted Endemism")

```

###Save the weighted endemism rasters and the images of it.

```{r}

outpath_tif_we <- file.path('data/out/wendemism_tif',c(paste0('wendemism_tif_',c(0.135, 0.225, 0.315, 0.405, 0.450),'.tif')))

outpath_png_we <- file.path('data/out/wendemism_png',c(paste0('wendemism_png_',c(0.135, 0.225, 0.315, 0.405, 0.450),'.png')))

titles_we <- c(paste0('Endemismo ponderado (escala ',c(0.135, 0.225, 0.315, 0.405, 0.450), '°)'))

lapply(seq_along(wendemism), function(x){
  save_raster(raster = wendemism[[x]], output_path =  outpath_tif_we[[x]])
  })

lapply(seq_along(wendemism), function(x){
  save_png(raster = wendemism[[x]], output_path =  outpath_png_we[[x]], title = titles_we[[x]])
  })

```

##Phylogenetic measures

*Here we upload the df of the tree and the records data frame*

```{r}
data_morton <- fread("../../data/in/filogenia/Hipp2019_sample_metadata.csv")
geo_data <- fread("../../data/in/hgbif_completo_iucn.csv")
```

Now, we view the similarities and the differencies between the two.

```{r}
#TODO: Revisar si los nombres en el correctname que no coinciden con los del morton realmnete no están en la filogenia 
#TODO: Decidir si esta es la filogenia que vamos a utilizar

common_morton <- data_morton[`Cleaned_NAMES-USE-THIS` %in% geo_data$correctname,]
common_geo <- geo_data[correctname %in% common_morton$`Cleaned_NAMES-USE-THIS`]

diff<- setdiff(unique(data_morton$`Cleaned_NAMES-USE-THIS`),unique(geo_data$correctname))

```

Existen 87 especies en común entre la filogenia y la base de datos de gbif/herbario.

### Tabla de especies comunes

To visualize how many of this species have less than 10 records, we graphed it:

```{r}
common_geo[,.N, by = correctname][order(-N), more_than_10 := N>10] %>% 
  ggplot(aes(y = fct_reorder(correctname, N), x = N, fill = more_than_10))+
  geom_col()+
  labs(x = "Numero de registros", y = "Especies")+
  theme_bw()

```

Now we save all the records in the gbif/herbario data frame that are present in the tree.
```{r}
common_geo <- merge(common_geo,unique(common_morton[,.(`Cleaned_NAMES-USE-THIS`,section,clade,subgenus)]), by.x = "correctname", by.y="Cleaned_NAMES-USE-THIS",all.x = T)

fwrite(common_geo, here::here('data/out/common_geo.csv'), row.names = F)
```


## Phylogenetic tree

### Clean names of the tree


The next function allows you to replace characters at the tips of the tree.

```{r}
library(ape)
library(treeio)
library(rotl)

## --- Functions --- ##

#' Modify Tip Labels of a Phylogenetic Tree
#'
#' This function modifies the tip labels of a phylogenetic tree based on a user-defined string transformation function.
#' By default, it removes substrings matching '_ott.*' and replaces underscores with spaces.
#'
#' @param tree A phylogenetic tree (likely of class `phylo`) containing tip labels.
#' @param modify_label_fn A function that takes a character vector of labels and returns a modified character vector.
#' The default function removes '_ott.*' and replaces underscores with spaces.
#'
#' @return A modified phylogenetic tree with updated tip labels.
#' @importFrom tibble as_tibble
#' @importFrom stringr str_replace str_remove str_detect
modify_tips <- function(tree, modify_label_fn = function(x) str_remove(x, '_ott.*')) {
  
  metadata <- as_tibble(tree) %>% mutate(label = modify_label_fn(label))
  tree$tip.label <- metadata[which(str_detect(metadata$label, 'Quercus')),]$label
  
  return(tree)
}

```

La filogenia [Hipp et al., 2019](https://nph.onlinelibrary.wiley.com/doi/10.1111/nph.16162) con largos de rama medidos en millones de años se puede extraer del Proyecto [Open Tree of Life](https://tree.opentreeoflife.org/curator/study/view/ot_1957/?tab=home). El paquete de `phyloraster` tiene dos opciones para calcular la diversidad filogenética. El primero es sumando los largos de rama de un subconjunto de la filogenia y el segundo es conservando los largos de rama pertenecientes a otros grupos no incluidos.

![Imagen tomada de la introducción al paquete [phyloraster](https://gabferreira.github.io/phyloraster/articles/phyloraster-vignette.html#references)](img/crop_tree_phyloraster.jpg)

Descargar el árbol de los encinos completo.

El arbol en crudo muestra etiquetas para cada especie. Como se muestra en el siguiente ejemplo: Quercus chrysolepis\|USA\|CA\|CA-DAV-MH20\|QUE000293'

Para este caso, no se utilizaran las etiquetas. Por lo que se utiliza la función modify_tips para eliminar las etiquetas y eliminar los espacios, por lo que el ejemplo quedaría de la siguiente manera: Quercus_chrysolepis

```{r}
complete_tree <- read.tree(here::here('data/in/filogenia/OpenTreeOfLife_hipp2019.tre'))


complete_tree <- modify_tips(complete_tree, modify_label_fn = function(x) str_replace_all(str_remove_all(str_remove(x, '\\|.*'),"\\'"),' |-','_'))

```

## Phylogenetic diversity calculation

Cargar las distribuciones de las especies.

Hacer un subset de los rasters que si se encuentran en el arbol (87).

```{r}
#files = rasters de distribucion por cada escala

sub_dist <- list()

for(i in seq_along(files)) {

names(files[[i]]) <- str_replace_all(names(files[[i]]),' |-','_')
rasters_to_use <- which(names(files[[i]]) %in% str_replace_all(unique(common_geo$correctname),' |-','_'))

sub_dist[[i]] <- subset(files[[i]], rasters_to_use)

}
```

```{r}
# Revisar que los nombres de las especies estén en el orden que están en el árbol

#Toma las puntas del arbol y busca las que no se encuentran en la base de datos, es decir, y con el sub_tree se eliminan
tips_to_drop <-  which(!complete_tree$tip.label %in% str_replace_all(unique(common_geo$correctname),' |-','_'))
sub_tree <- drop.tip(complete_tree,tips_to_drop)


# Con el paquete phyloraster, se reacomodan los datos de los rasters para que esten en el orden de la filogenia. 

#Esto arroja un Warning, el cual indica que las especies Quercus_depressipes, Quercus_gambelii no se encuentran en el arbol. Por lo que solo se estaria calculando la PD de 85 especies.

dataprep <- list()
pdr <- list()

for(i in seq_along(sub_dist)) {
  
  #tabla con especies ordenadas
dataprep[[i]] <- phylo.pres(x = sub_dist[[i]], tree = sub_tree)

  #el raster del calculo
pdr[[i]] <- rast.pd(x = dataprep[[i]]$x, dataprep[[i]]$tree)

}
```

###Save the PD rasters and the images.

Guardar los rasters con el calculo y las graficas en formato .png

```{r}


outpath_tif_pd <- file.path('data/out/PD_tif',c(paste0('PD_tif_',c(0.135, 0.225, 0.315, 0.405, 0.450),'.tif')))

outpath_png_pd <- file.path('data/out/PD_png',c(paste0('PD_png_',c(0.135, 0.225, 0.315, 0.405, 0.450),'.png')))

titles_pd <- c(paste0('Diversidad Filogénetica de especies (escala ',c(0.135, 0.225, 0.315, 0.405, 0.450), '°)'))

lapply(seq_along(pdr), function(x){
  save_raster(raster = pdr[[x]], output_path =  outpath_tif_pd[[x]])
  })

lapply(seq_along(pdr), function(x){
  save_png(raster = pdr[[x]], output_path =  outpath_png_pd[[x]], title = titles_pd[[x]])
  })

```

##Richness of the species present in the tree

```{r}

richness_tree <- lapply(sub_dist, rast.sr)
```

###Save the richness rasters and the images of it.

```{r}

outpath_tif_rich_tree <- file.path('data/out/richness_tif_tree_species',c(paste0('richness_tif_',c(0.135, 0.225, 0.315, 0.405, 0.450),'.tif')))

outpath_png_rich_tree <- file.path('data/out/richness_png_tree_species',c(paste0('richness_png_',c(0.135, 0.225, 0.315, 0.405, 0.450),'.png')))

titles_richness_tree <- c(paste0('Riqueza de especies presentes en el árbol filogénetico (escala ',c(0.135, 0.225, 0.315, 0.405, 0.450), '°)'))

lapply(seq_along(richness_tree), function(x){
  save_raster(raster = richness_tree[[x]], output_path =  outpath_tif_rich_tree[[x]])
  })

lapply(seq_along(richness_tree), function(x){
  save_png(raster = richness_tree[[x]], 
          output_path =  outpath_png_rich_tree[[x]], 
          title = titles_richness_tree[[x]])
  })

```


##Weighted endemism of the species present in the tree

```{r}

wendemism_tree <- list()


for(i in seq_along(files)) {

  
wendemism_tree[[i]] <- rast.we(x = sub_dist[[i]])

}

plot(wendemism_tree[[5]]$WE, main = "Weighted Endemism")

```

###Save the weighted endemism rasters and the images of it.

```{r}

outpath_tif_we_tree <- file.path('data/out/wendemism_tif_tree',c(paste0('wendemism_tif_',c(0.135, 0.225, 0.315, 0.405, 0.450),'.tif')))

outpath_png_we_tree <- file.path('data/out/wendemism_png_tree',c(paste0('wendemism_png_',c(0.135, 0.225, 0.315, 0.405, 0.450),'.png')))

titles_we_tree <- c(paste0('Endemismo ponderado para especies presentes en el árbol filogénetico (escala ',c(0.135, 0.225, 0.315, 0.405, 0.450), '°)'))


lapply(seq_along(wendemism_tree), function(x){
  save_raster(raster = wendemism_tree[[x]], output_path =  outpath_tif_we_tree[[x]])
  })



lapply(seq_along(wendemism_tree), function(x){
  save_png(raster = wendemism_tree[[x]], output_path =  outpath_png_we_tree[[x]], title = titles_we_tree[[x]])
  })



```

