---
title: "extract_cleaning"
author: "Gabriela Centeno y Sofía Zorrilla"
format: html
knitr:
  opts_knit:
    root.dir: "../../"
---

This .qmd documents obtains the elevation for each one of the registers. We use the raster called "Continuo de Elevacion Mexicano (CEM)" from INEGI. The resolution is 15 m x 15 m.

First, upload the neccesary libraries.
```{r}
library(sf)
library(data.table)
library(here)
library(terra)
library(elevatr)
```

-Upload the data frame with the registers.
```{r}
hgbif_clean <- fread(here::here("data/in/hgbif_clean.csv"))
```


```{r}

#Genera el df de elevacion

# Reordena las columnas de la tabla para que decimalLongitude sea la primera y decimalLatitude la segunda

hgbif_clean_coords <- hgbif_clean %>%
  select(-starts_with("V1")) %>%
  mutate(
    decimalLongitude = as.numeric(decimalLongitude),
    decimalLatitude = as.numeric(decimalLatitude)
  ) %>%
  select(decimalLongitude, decimalLatitude)  %>%
  filter(decimalLongitude >= -180 & decimalLongitude <= 180, decimalLatitude >= -90 & decimalLatitude <= 90)

#Convert it into an sf object 

hgbif_clean_sf <- st_as_sf(hgbif_clean,coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)

```

```{r}

# Obtener elevación para cada punto en tu base de datos
elev_data <- get_elev_point(hgbif_clean_sf, 
                            src = "aws", # Usa Amazon Web Services como fuente de datos de elevación
                            z = 7)      # Nivel de zoom (determina la precisión) 7 parece ser 1 km (https://github.com/tilezen/joerd/blob/master/docs/data-sources.md#what-is-the-ground-resolution)


#guardar en un csv 
write.csv(elev_data, here::here("data/in/elevation/elev_data.csv"))

```


```{r}

# Set file paths
raster_path <- here::here("data/in/CEM/CEM_15m_ITRF08.tif")  # Path to the large TIFF file
output_csv <- here::here("data/in/elevation/elevation_INEGI.csv")     # Output CSV file

# Load the raster efficiently
raster <- rast(raster_path)


# Ensure CRS matches
if (!st_crs(hgbif_clean_sf) == crs(raster)) {
  points_sf <- st_transform(hgbif_clean_sf, crs(raster))
}

# Convert sf object to SpatVector for efficient extraction
points_vect <- vect(points_sf)

# Extract elevation values
extracted_values <- terra::extract(raster, points_vect)

# Merge extracted values with the original sf object
points_sf$elevation <- extracted_values[, 2]  # Assuming first column is ID

# Write to CSV
st_drop_geometry(points_sf) |> 
  write.csv(output_csv, row.names = FALSE)

```

-Generate a data frame that has coordinates that can be mapped 
```{r}
hgbif_clean_elev <- hgbif_clean %>%
  select(-starts_with("V1")) %>%
  mutate(
    decimalLongitude = as.numeric(decimalLongitude),
    decimalLatitude = as.numeric(decimalLatitude)
  ) %>%
  select(decimalLongitude, decimalLatitude,everything())
```

-Add the elevation to the data frame we just created.
```{r}
#Agregar columna de elevacion
elev_data_elevatr <- fread(here::here("data/in/elevation/elev_data.csv"))

elev_data_inegi <-
  fread(here::here("data/in/elevation/elevation_INEGI.csv"))

hgbif_clean_elev$elevacion_elevatr <- elev_data_elevatr$elevation

#Add INEGI elevation 
hgbif_clean_elev$elevacion_inegi <-
  elev_data_inegi$elevation

#Filter the registers were the elevation is equal or more than 0.
hgbif_clean_elev3 <- hgbif_clean_elev %>%
  filter(elevacion_inegi >= 0)

#Save it in a csv archive
write.csv(hgbif_clean_elev, here::here("data/in/hgbif_clean_elev.csv"))

```


-Proceed to location_extraction_iucn_add.qmd