---
title: "location_extraction"
author: "Gabriela Centeno y Sofía Zorrilla"
format: html
knitr:
  opts_knit:
    root.dir: "../../"
---

This .qmd document extract the state and municipality for each one of the registers.

First, upload the necessary libraries.
```{r}
library(data.table)
library(tidyverse)
library(sf)
```


Upload the data frame with the registers, coordinates and elevations and the layer that has the location information. In this case we use a shape file that contains all the municipalities from CONABIO (2022).

```{r}
#Subir las capas 
hgbif_clean_elev <- fread(here::here("data/in/hgbif_clean_elev.csv"))

mun <- st_read(here::here("data/in/mun22gw/mun22gw.shp"))

```

-Then we select the columns from the shape object of interest, in this case is the name of the municipality, the state and the geometry.
```{r}
# Select only the relevant columns from the sf objects
mun2<- mun[, c("NOMGEO", "NOM_ENT", "geometry")]
```

-Then convert the data frame that has the registers into a sf object, so we can join this data frame with the mun22gw.shp.
```{r}
points_hgbif <- st_as_sf(hgbif_clean_elev, coords = c("decimalLongitude", "decimalLatitude"), crs = 4326) 

points_mun_state <- st_join(points_hgbif, mun2)

coordinates <- st_coordinates(points_mun_state)

points_mun_state <- cbind(points_mun_state, coordinates)

hgbif_completo <- as.data.frame(points_mun_state) %>% 
  select(-1, -2, -"stateProvince", -"county",-"municipality") 

#Save it in a csv
write.csv(hgbif_completo, here::here("data/in/hgbif_elev_mun.csv"))

```

-Union of metadata from state and municipality with the IUCN conservation status. We obtained the conservation status using the script extract_IUCN.R.

```{r}

hgbif_completo <- fread(here::here("data/in/hgbif_elev_mun.csv"))

iucn <- fread(here::here("data/in/conservation_status/IUCN_per_species.csv"))

#Check that the names are equal.
setdiff(hgbif_completo$correctname, iucn$correctname)
setdiff(iucn$correctname, hgbif_completo$correctname)

# Join tables
hgbif_completo2 <- merge(hgbif_completo, iucn[,-2], by = "correctname")

# Remove unnecessary columns
hgbif_completo2 <- hgbif_completo[, !c("V1","V20","geometry", "elevacion")]

hgbif_completo2_clean <- hgbif_completo2_clean %>%
   filter(!is.na(X) & !is.na(Y)) %>%
    filter(X >= -118 & X <= -87,  
           Y >= 14 & Y <= 33) 
  
#Write in a csv
fwrite(hgbif_completo2_clean, here::here("data/in/hgbif_completo_iucn.csv"), row.names = F)
```
