---
title: "richness_wendemism_calculation"
author: "Gabriela Centeno y Sofía Zorrilla"
format: html
eval: false
knitr:
  opts_knit:
    root.dir: "../../"
---

*This .qmd document calculates the richness and weighted endemism for 213 species using the phyloraster package.*

*Then, it calculates the richness and weighted endemism for the 105 species present in the phylogenetic tree.*

*Next, it does the same thing for each section (Lobatae, Quercus and Virentes).*

-   Load the necessary libraries

```{r}
library(phyloraster)
library(terra)
library(ape)
library(phylobase)
library(here)
library(data.table)
library(here)
```

-   Enter the stack of rasters of the absence/presence of the 213 species we made previously. Then calculate the richness with function rast.sr, finally plot it.

```{r}
ras <- terra::rast(here::here("data/out/raster_species/ras_species_res_0.5.tif"))
sr <- rast.sr(x = ras)

plot(sr, main = "Species richness")
```

-   Save the richness raster and an image of it.

```{r}
#| eval: false 

#Save it
tiff(here::here("data/out/phyloraster_maps/species_richness.tiff"), width = 3000, height = 2000, res = 300, compression = "lzw")
plot(sr, main = "Species richness")
dev.off()

#Save it in png
png(here::here("data/out/phyloraster_maps/species_richness.png"), width = 3000, height = 2000, res = 300)
plot(sr, main = "Species richness")
dev.off()
```

-   Calculate the weighted endemism with the function rast.we, then plot it.

```{r}
wer <- rast.we(x = ras)
plot(wer$WE, main = "Weighted Endemism")
```

-   Save the weighted endemism raster and an image of it.

```{r}
#Save it in tiff
tiff(here::here("data/out/phyloraster_maps/species_WEndemism.tiff"), width = 3000, height = 2000, res = 300, compression = "lzw")
plot(wer$WE, main ="Weigthed Endemism")
dev.off()

#Save it in png
png(here::here("data/out/phyloraster_maps/species_WEndemism.png"), width = 3000, height = 2000, res = 300)
plot(wer$WE, main ="Weigthed Endemism")
dev.off()
```

-   Make the stack of rasters of the absence/presence of the 105 species present in the phylogenetic tree.

-   To do this, we need to filter the species that are not present in the tree. To do this we need to load the data frame we generated before with all the species that are in the tree. Then, we proceed to subset the raster with this information.

```{r}
common_geo <- fread(here::here("data/out/common_geo.csv"))

dist <- rast(here::here('data/out/raster_species/ras_species_res_0.5.tif'))

rasters_to_use <- which(names(dist) %in% unique(common_geo$correctname))

sub_rast <- subset(dist,rasters_to_use)
```

-   Calculate the richness and weighted endemism for the species present in the tree. 

```{r}
sr_tree <- rast.sr(x = sub_rast,here::here("data/out/phyloraster_maps/tree/richness_tree_species_res_0.5.tif"))

plot(sr_tree, main = "Species richness_subset")

wer_tree <- rast.we(x = sub_rast,here::here("data/out/phyloraster_maps/tree/wendemism_tree_species_res_0.5.tif"))

plot(wer_tree$WE, main = "Weighted Endemism")

```

-   Next, we will calculate the same for each section *(Lobatae, Quercus and Virentes)*.

-   First, we will make a subset of the stack of raster previously made. So we will have three new stacks of rasters, one for each section. 
-   We will follow the same procedure than before, but this time we are going to specify to filter the registers of interest. 

```{r}


rasters_to_use <- which(names(dist) %in% unique(common_geo$correctname))

sub_rast <- subset(dist,rasters_to_use)
```
